Biden Expected to Win 54% of Popular Vote in 2020 US Presidential Election

### 2020-11-02 outputs/paper/paper.Rmd
updated model discussion

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Our model is built on MRP, which stands for multi-level regression with post-stratifications. We utlize multi-level regression in order to create a model that we can build predictions upon. Since our predictor y is a binary variable (Biden or Trump), we used a logistic regression in order build our model. A logistic regression predicts the likelihood of an outcome based on various independent variables and it fits a model on a log curve which is bound in the Y axis in the range of [0,1].. In our case, we judge the likelihood that an individual would vote for Joe Biden in comparison to Donald Trump, given the individuals age, gender, education, state, and race. Given these inputs and outputs, we can train our logistic regression model using BRMS to predict which values for each variable predicts certain votes, and this will be discussed further in the later section. In our model, as it is a binary predictor, we round the prediction to the nearest number (more than 0.5 is rounded to 1, meaning this individual is likely to vote for Joe Biden, and less than 0.5 is rounded to 0, meaning this individual is likely to vote to Donald Trump).

One variable we discussed about including was annual income, however when we added the income lowered our cross validation score and the well discussed issues with self-reported income (bias, inaccurate estimations, not reporting) convinced us to disclude income. Another variable we debated on was education. Originally, education was included, however, education severely impacted our cross validation accuracy. We solved these issues by summarizing all education levels into 2 levels: at least some Post-Secondary and No post-secondary. This allowed our model to perform slightly better, but more importantly, there was a significant difference in the voting preference between the 2 groups. Another variable that helped increase accuracy after manipulation was age, where it became a better independent variable after breaking age into groups on 10's (eg. 18-28, 29-38) in order to better represent different age groups. Other groupings we tried was breaking into youth, middle age, and seniors (18-35, 36-55, 55+) and leaving age as individual numeric values, however both of these breakdowns did not provide any significant value.

Overall, our model performs fairly (Figure N), with a cross-validation accuracy of approximately 63% (Figure N). This performed better than 50% (guessing) and through this we are able to gain some insight on how each variable affect the prediction. Furthermore, by investigating our inaccuracies (Figure N), we see that of the approximately 21% of our predictions were false negatives (Predicted Trump vote when it should predict Biden) and 15% of our predictions were false positives (predicted Biden when it should have predicted Trump vote). This was significantly better than some of the other models we tried as discussed in the previous paragraphs, as we had a large number of false negatives in many models when education was split into multiple groups rather than just College or no College.

@@

# References
R
Survey and ACS dataset
https://www.voterstudygroup.org/uploads/reports/Data/Nationscape-User-Guide_2020sep10.pdf
https://www.voterstudygroup.org/uploads/reports/Data/NS-Methodology-Representativeness-Assessment.pdf
Wu
https://www.pewresearch.org/fact-tank/2020/10/26/what-the-2020-electorate-looks-like-by-party-race-and-ethnicity-age-education-and-religion/
ACS stuff:
https://www2.census.gov/geo/pdfs/education/Uhl_CAS_2011.pdf
https://www.census.gov/programs-surveys/acs/technical-documentation/errata/120.html
https://usa.ipums.org/usa/resources/voliii/formACS2018.pdf
https://usa.ipums.org/usa/resources/codebooks/ACS_codebook.pdf
https://www.politifact.com/factchecks/2014/jan/09/us-census-bureau/americans-must-answer-us-census-bureau-survey-law-/

>>>>

The purpose of our model is to predict a person's vote in the 2020 US election given their characteristics (eg. income, race, gender, etc.). The dependent variable is a binary categorical variable, where 0 represents voting for Trump and 1 represents voting for Biden. The independent variables are the characteristics of the person. To avoid multicollinearity which causes unreliable regression estimates, a few carefully selected independent variables were used for our model. One criteria for the selection was that the variables had to be both present in the Nationscapes dataset and the ACS dataset, so that they be included for post-stratification. Another criteria is that the characteristic should group people that share a similar perspective which may impact on their voting decision. Through literature research (Uppal and LaRochelle-Cote, 2015), we found that some helpful indicators were age, gender, education, state, race, and self-reported income. To finalize the selection, characteristics should not be highly correlated and the data provided should make logical sense. Since the data quality for income was generally poor and it worsened the model performance, income was included in the model. Therefore, the predictors used for the model are age, gender, education, state, and race.

Since the goal is to do binary classification using multiple explanatory variables, binary logistic regression is a suitable method for the model. We used logistic regression to predict the probability of a person's voting choice and determine their vote by rounding to the nearest represented binary categorical variable. Logistic regression takes independent variables as inputs and based on the assigned weights and a logit function, the output will be a probability [0,1] of the outcome happening. Below is the equation of the logistic regression model:

\begin{equation}
Pr(Y_i \in \{Trump, Biden\}) = logit^{-1}(\alpha_{j[i]}^{\textrm{age}} + \alpha_{j[i]}^{\textrm{gender}} + \alpha_{j[i]}^{\textrm{edu}} + \alpha_{j[i]}^{\textrm{state}} + \alpha_{j[i]}^{\textrm{race}})
\end{equation}

The Nationsacpes dataset contains the characteristics of a voter and also their vote, so the model was trained by using cross validation split on that dataset. 95% of the dataset was used as the training set to determine the best weights for the model and the remaining 5% of the dataset was used for the test set to verify the accuracy of the model. The vast majority of the data is used for the training set because we wanted to ensure that the data does not overfit due to lack of data. The logistic regression model was implemented using the BRMS library.

Furthermore, there were many adjustments that had to be done with the inputs to improve model accuracy. Apart from self-reported income, education was another independent variable that was debated on whether it should be included in the model. Education severely impacted our cross validation accuracy, due to a wide range of values which were too specific for our purpose. These issues were resolved by summarizing all education levels into 2 levels: At least some Post-Secondary and No post-secondary. This allowed our model to perform slightly better, but more importantly, there was a significant difference in the voting preference between the 2 groups.

Another variable that initially caused inaccurate results was age. There was no clear trend in having age as a numeric continuous value for predicting votes. As a result, the numerical age values were grouped into 10's (eg. 18-28, 29-38) in order to better represent different age groups. Other groupings we tried was breaking into youth, middle age, and seniors (18-35, 36-55, 55+); however, that split was not able to capture the pattern in the age groups because the model accuracy worsened. So grouping age by 10's was the adjustment that gave better results and this is the configuration that was included in the model.

Overall, our model performs fairly (Figure N), with a cross-validation accuracy of approximately 63% (Figure N). This performed better than 50% (guessing) and through this we are able to gain some insight on how each variable affect the prediction. Furthermore, by investigating our inaccuracies (Figure N), we see that of the approximately 21% of our predictions were false negatives (Predicted Trump vote when it should predict Biden) and 15% of our predictions were false positives (predicted Biden when it should have predicted Trump vote). This was significantly better than some of the other models we tried as discussed in the previously, as we had a large number of false negatives in many models when education was split into multiple groups rather than just College or no College.

@@

# References
R
Survey and ACS dataset
https://www.voterstudygroup.org/uploads/reports/Data/Nationscape-User-Guide_2020sep10.pdf
https://www.voterstudygroup.org/uploads/reports/Data/NS-Methodology-Representativeness-Assessment.pdf
Wu
https://www.pewresearch.org/fact-tank/2020/10/26/what-the-2020-electorate-looks-like-by-party-race-and-ethnicity-age-education-and-religion/

Model Stuff:
https://www150.statcan.gc.ca/n1/pub/75-001-x/2012001/article/11629-eng.htm

ACS stuff:
https://www2.census.gov/geo/pdfs/education/Uhl_CAS_2011.pdf
https://www.census.gov/programs-surveys/acs/technical-documentation/errata/120.html
https://usa.ipums.org/usa/resources/voliii/formACS2018.pdf
https://usa.ipums.org/usa/resources/codebooks/ACS_codebook.pdf
https://www.politifact.com/factchecks/2014/jan/09/us-census-bureau/americans-must-answer-us-census-bureau-survey-law-/
